DeepSphere: Efficient spherical Convolutional Neural Network with HEALPix sampling for cosmological applications

### 2018-10-23 scnn.tex
removed comments from latex

%Convolutional Neural Networks (CNNs) are becoming an important analysis tool in cosmology and astrophysics.
% MD: useful info? To me this is not an argument for our approach. Maybe we have to expand it?

% Architectures are then designed similarly to the classical CNNs used for image analysis.
% HealpixNet uses Tensorflow as the underlying engine and can utilize most of it functionalities.
% MD: not relevant for the abstract in my opinion

Convolutional Neural Networks (CNNs) are a cornerstone of the Deep Learning toolbox and have led to many breakthroughs in Artificial Intelligence.
So far, these networks have mostly been developed for regular Euclidean domains such as those supporting images, audio, or video.
Because of their success, CNN-based methods are becoming increasingly popular in Cosmology.
Cosmological data often comes as spherical maps, which make the use of traditional CNNs more complicated.
The commonly used pixelization scheme for spherical maps is the Hierarchical Equal Area isoLatitude Pixelisation (HEALPix).
We present a spherical CNN for analysis of full and partial HEALPix maps, which we call DeepSphere.
The spherical CNN is constructed by representing the sphere as a graph.
Graphs are versatile data structures that can represent pairwise relationships between objects or act as a discrete representation of a continuous manifold.
Using the graph-based representation, we define many of the standard CNN operations, such as convolution and pooling.
With filters restricted to being radial, our convolutions are equivariant to rotation on the sphere, and DeepSphere can be made invariant or equivariant to rotation.
This way, DeepSphere is a special case of a graph CNN, tailored to the HEALPix sampling of the sphere.
% TBD if we include it
This approach is computationally more efficient than using spherical harmonics to perform convolutions.
% An advantage of the graph representation is that the convolution operations are efficient, which enables the CNN to run fast.
We demonstrate the method on a classification problem of mass maps from two cosmological models and compare the performance of the CNN with that of two baseline classifiers, based on the power spectrum and pixel density histogram.
Our experimental results show that the performance of DeepSphere is always superior or equal to both of these baselines.
For high noise levels and for data covering only a smaller fraction of the sphere, DeepSphere achieves typically 10\% better classification accuracy than those baselines.
Finally, we show how learned filters can be visualized to introspect the neural network.
Code and examples are available at \url{https://github.com/SwissDataScienceCenter/DeepSphere}.

%As a response to that fact, we are witnessing an increasing interest in the generalization of CNNs to graphs.
% Graphs are versatile data structures which can represent pairwise relationships between objects or act as a discrete representation of a continuous manifold.
% In this paper we take the latter view and represent a sampled sphere as a graph, for which we define the convolution and pooling operations.
% Spherical maps appear in many applications, such as the study of phenomena on the Earth for meteorologists or on the sky for cosmologists.
% A popular format for spherical maps is HEALPix, which uses an equal-area, isolatitude sampling.
% Hence, we present a spherical CNN for analysis of full and partial HEALPix maps, which we call DeepSphere.
% \nati{I am not that we should talk about pooling in the abstract. Another angle would be to say that HEALPix sampling is well suited for spherical CNN with graphs as a) it is very regular => the graph capture well the spherical structure of the Manifold and b) pooling is natural.}
% While the convolution on arbitrary graphs has mostly been sorted out, a general pooling operation has not been discovered so far.
% As such, while DeepSphere is a special case of a graph CNN, it is tailored to the HEALPix sampling of the sphere.
% This approach is computationally much more efficient than using spherical harmonics to perform convolutions, an alternative that has been explored previously.
% The filters learned by the DeepSphere are restricted to being radial.
% Michaël: not necessary in the abstract?
% We demonstrate the method on a classification problem of mass maps from two cosmological models.
% \nati{Should we specify what are the baselines? I think not.}
% A standard benchmark for that problem is to train an SVM on the histograms of map pixel values. Another is to use the power spectral density instead.
% DeepSphere achieves \todo{xx\%} better accuracy than those baselines, is more robust to noise, and is \todo{faster?}.
% Code and examples are available at \todo{\url{https://github.com/xxx}}.
\end{abstract}

@@

% \section*{TODO list}
% \begin{itemize}[noitemsep,topsep=0pt,parsep=0pt,partopsep=0pt]
% 	\item (TK) Can we use HEALPix in the name? Check with the main author of HEALPix. MD: what about DeepCosmo or CosmoNet? NP: cosmological data is not necessarily spherical.
% 	\nati{No answer from the people. Let us go for DeepSphere. Graphsphere???}
% 	\item (TK) Check use of British vs American English. Check use of tense.
% 	\item (MD) Figure 4: speed for multiple lmax. Not doing it.
%     \item (TK) check part sphere notebook and whole sphere
% 	\item (NP \& MD) Move the git repository, rename the classes, check the links.
%     \item (Everyone should check this) Notation: write element $i$ of vector $y$ as $y_i$, $y(i)$, or $y[i]$? Be consistent. Same for matrices. Keep indices.
% 	\item (fail - I checked few fonts here: \url{de.sharelatex.com/learn/Font_typefaces}, but none looked better) (TK) Font: do we have to use times (set in the documentclass)? I (MD) find the rendering of math a bit ugly. MD: lmodern is better, but can we use it (w.r.t.\ journal policy)? If equations split out, we can try to arrange them. We should also check which font size we can / have to use.
% \end{itemize}

\begin{figure*}
	\centering
    % \includegraphics[width=\linewidth]{architecture}
	\includegraphics[width=\linewidth]{figure_architecture_v3}
	\caption{Overall network architecture, showing here three convolutional layers acting as feature extractors followed by three fully connected layers with softmax acting as the classifier.
    A convolutional layer is based on five operations: convolution, non-linearity, batch normalization, down-sampling, and pooling.
	While most operations are agnostic to the data domain, the convolution and the down-sampling have to be adapted.
	In this paper, we propose first to model the sphere with a graph and to perform the convolution on the graph.
	Graphs are versatile data structures which can model any sampling, even irregular or partial.
	Second, we propose to exploit a hierarchical pixelization of the sphere for the down-sampling operation.
	It allows the network to analyze the data at multiple scales while preserving the spatial localization of features.
	This figure shows a network that operates on the whole sphere.
	The process is the same when working with partial observations, except that the graph is only built for the region of interest.}
	\label{fig:architecture}
\end{figure*}

@@

% Structure:
  % High-level intro
  % Desiderata for a method
  % Related methods / previous work
  % Our method
  % Experiments

@@

% cohen2017convolutional (ICML workshop) and cohen2018sphericalcnn (ICLR best paper) are redundant
One direction to generalize CNNs to spherical data is to apply a standard 2D CNN to a grid discretisation of the sphere~\citep{boomsma2017spherical,su2017sphericalconv,coors2018spherenet}.
An alternative is to divide the sphere into small chunks and project those on flat 2D surfaces \citep{fluri2018deep,gupta2018nongaussianinformation,schmelze2017cosmologicalmodel,gillet2018deeplearning}.
While these approaches use the well-developed 2D convolution and hierarchical pooling, they are not equivariant to rotation.
Another direction is to leverage the spherical Fourier transform and to perform the convolution associated to the SO(3) rotation group in the spectral domain, thanks to the convolution theorem \citep{cohen2018sphericalcnn, esteves2017sphericalcnn}.
While the resulting convolution is equivariant to rotation, this approach is computationally expensive, even if a fast spherical Fourier transform is used.
Moreover, all those methods cannot be much accelerated when maps only span a part of the sky.

% MD: I don't see what the following brings to the paper.
% This work is the first to design a spherical convolutional neural network for cosmological applications.
% Previously, several other machine learning algorithms were extended to the sphere, often also including the third, radial direction.
% Most notably, sparsity-based methods were developed, such as wavelets \citep{Leistedt2013flaglets,Leistedt2015spinwavelets}, ridgelets \citep{McEwen2015ridgelet} and curvelets \citep{Chan2017curvelets,Starck2009polarized}.

Our spherical CNN leverages convolutions on graphs and hierarchical pooling to address all three identified shortcomings: (i) rotation equivariance, (ii) computational efficiency, and (iii) partial sky observations.
The main idea is to model the discretised sphere as a graph of connected pixels: the length of the shortest path between two pixels is an approximation of the geodesic distance between them.
This approach is enabled by recent developments in graph signal processing, where fundamental tools, such as the Fourier transform, were defined for graphs \citep{shuman2013emerging}.
The graph convolution has been explored previously for omnidirectional imaging \citep{khasanova2017graphomni}, with a different down-sampling operation.
Our pooling strategy exploits a hierarchical pixelisation of the sphere to analyse the data at multiple scales.
% Compared with~\citep{khasanova2017graphomni}, our work does not need hand-chosen statistics nor dynamic pooling.
We use the graph CNN formulation introduced in \citep{defferrard2016convolutional}, which can accommodate any sampling of the sphere.
As the Equal Area isoLatitude Pixelisation (HEALPix)~\citep{gorski2005healpix} is a popular sampling used in cosmology and astrophysics, we tailored the method to that particular sampling and call it DeepSphere.
DeepSphere is however easily used with other samplings as only two elements depend on it: (i) the choice of neighboring pixels when building the graph, and (ii) the choice of parent pixels when building the hierarchy.
The flexibility of modeling the data domain with a graph allows to easily model data that spans only a part of the sphere, or data that is not uniformly sampled.
Using a $k$-nearest neighbours graph, the convolution operation costs $\mathcal{O}(N_{pix})$ operations, where $N_{pix}$ is the number of pixels.
This is the lowest possible complexity for a convolution without approximations.
DeepSphere is readily apt to solve four tasks: (i) global classification (i.e., predict a class from a map), (ii) global regression (i.e., predict a set of parameters from a map), (iii) dense classification (i.e., predict a class for each pixel of a map), and (iv) dense regression, (i.e., predict a set of maps from a map).
Input data are spherical maps with a single value per pixel, such as the CMB temperature, or multiple values per pixel, such as surveys at multiple radio frequencies.

%\nati{I am not sure this goes there...} Other recently developed sampling methods include \citep{mcewen2011novelsampling}, where exact sampling theorem for equiangular MW sampling was presented.
We give a practical demonstration of DeepSphere on cosmological model discrimination using maps of projected mass distribution on the sky \citep{chang2017curvedsky}.
This kind of maps can be created using the gravitational lensing technique \citep[see][for review]{BartelmannSchneider2001weak}.
Our maps are similar to the ones used by \citep{schmelze2017cosmologicalmodel}.
In a simplified scenario, we classify partial convergence maps into two cosmological models.
These models were designed to have very similar angular power spectrum.
We compare the performance of DeepSphere to a baseline algorithm: an SVM classifier that takes pixel histograms or power spectrum densities (PSDs) of these maps as input.
The comparison is made as a function of the additive noise level and the area of the sphere used in the analysis.
Results show that our model is always better at discriminating the maps, especially in the presence of noise.
DeepSphere is implemented with \pkg{TensorFlow} \citep{abadi2016tensorflow} and is intended to be easy to use out-of-the-box for cosmological applications.
Code and examples are available online.{\interfootnotelinepenalty=10000 \footnote{\url{https://github.com/SwissDataScienceCenter/DeepSphere/}}}

% MD: I usually find those paragraphs boring and skip them when reading. Our intro is long enough already. Feel free to add it back if you think it's necessary.
%The paper is organized as follows.
%In \secref{related_work}, we describe existing approaches to convolutions on spheres, graphs, and manifolds.
%\secref{method} exposes the construction of a graph given the HEALPix sampling of a sphere and defines the convolution operation.
%Finally, \secref{experiments} presents our experimental results on the weak lensing mass map classification problem. There we compare our method to two standard benchmarks and show its superiority.
%We conclude in \secref{conclusion}.

% Cite the following papers:
% \begin{itemize}
%     \item HEALPix paper \citep{gorski2005healpix}
%     \item First DES mass maps \citep{chang2017curvedsky}
%     \item HEALPix convolutions with asymetric beams \citep{mitra2011fastpixel}
%     \item Mass mapping on the sphere \citep{wallis2017mappingdark}
%     \item Wavelets on the sphere \citep{leistedt2016wavelet} and ball \citep{leistedt2012exactwavelets}
%     \item Planck main result \citep{planck2015cosmologicalparameters,komatsu2011sevenyear}
%     \item HI map \citep{HI4PI2016fullskyHI,}
%     \item more radio maps from Adam+Hamsa \citep{santos2015cosmologySKA}
%     \item more astro science on the sphere, what other survey make maps?
%     \item alternative sampling on the sphere \citep{mcewen2011novelsampling} MW sampling
%     \item cite kids tomographic power spectrum \citep{koehlinger2017kidstomographic}
% \end{itemize}

% \subsection{Potential applications}
%   \assign{Tomek, Nathanaël, Michaël}

% The analysis of spherical cosmological data, such as the cosmic microwave background \cite{...}, as done in \cite{he2018analysis}, is the target application of our method.

% While our method was developed with cosmology in mind, it can easily target any problem where data live on a sphere. Examples include, but are certainly not limited to, (i) efficient compression and decompression of \ang{360} videos (see \cite{su2017learning}), (ii) \todo{data analysis on planets? (climate, forecasting, temperature, wind)}, (iii) \todo{particle physics? (jets on detectors, but they are usually cylindrical)}, (iv) \todo{applications in Cohen's papers?}.

% Finally, not that those neural networks are not restricted to the sphere and can be applied to any problem where we have data on a graph, such as social, biological or infrastructure networks [some citations, e.g. brain Alzeihmer, particle physics, computer graphics].
% the convolution is not restricted to the sphere, the coarsening/pooling is

\section{Method}
\label{sec:method}
% \begin{itemize}
%   \item We build a graph using the healpix sampling
%   \item Define Fourier transform and show that the harmonics are visually close to the spherical harmonics
%   \item Define spherical convolution using the graph Fourier transform and show heat diffusion example
%   \item Show the limits of the approach and explain why we cannot have a perfect spherical convolution with this technique
% \end{itemize}

@@

% \begin{figure}
%     \centering
%     \begin{subfigure}[b]{0.45\linewidth}
%         \centering
%         \includegraphics[width=\linewidth]{exampleEarthTopo}
%         \caption{}
%         \label{fig:example_earth}
%     \end{subfigure}
%     \hfill
%     \begin{subfigure}[b]{0.45\linewidth}
%         \centering
%         \includegraphics[width=\linewidth]{exampleCMB}
%         \caption{}
%         \label{fig:example_cmb}
%     \end{subfigure}
%     \caption[]{Example of (a) an Earth topography map composed of \textasciitilde 3 million pixels (\textasciitilde 7 arcmin resolution) and (b) a model of the Cosmic Microwave Background (CMB) radiation temperature anisotropy, composed of \textasciitilde 13 million pixels (\textasciitilde 3.4 arcmin resolution).\footnotemark}
%     \label{fig:example_maps}
% \end{figure}

Before doing any numerical analysis on the sphere, one first has to choose a tessellation, i.e., an exhaustive partition of the sphere into finite area elements, where the data under study is quantized.
The simplicity of the spherical form belies the intricacy of global analysis on the sphere.
There is no known point set that achieves the analogue of uniform sampling in Euclidean space and allows exact and invertible discrete spherical harmonic decompositions of arbitrary but band-limited functions.
Any existing proposition of practical schemes for the discrete treatment of such functions on the sphere introduces some systematic error dependent on the global properties of the point set.
While our method is applicable to any pixelisation of the sphere, two details depend on the chosen sampling: (i) the choice of neighbours in the construction of the graph, and (ii) the choice of parent nodes when coarsening the graph.
As HEALPix~\citep{gorski2005healpix} is our target application, we tailor the method to that particular sampling in the subsequent exposition.
\figref{example_maps} shows three examples of HEALPix maps: the Cosmic Microwave Background \citep{planck2015overview}, galaxies found in Sloan Digital Sky Survey Data Release 14 \citep{abolfathi2017sdssDR14}, and an example simulated mass map on the footprint of Dark Energy Survey Data Release 1 \citep{des2018dr1}.
%We'll note in later sections where the method should be adapted for a different sampling.

HEALPix is a particular case of a more general class of schemes based on a hierarchical subdivision of a base polyhedron.
%Many discretized maps of the sphere are based on a hierarchical subdivision of a base polyhedron.
Another example is the geodesic grids which are based on geodesic polyhedrons, i.e., polyhedrons made of triangular faces. A counter-example is the equirectangular projection, which is not constructed from a base polyhedron, although it can be subdivided.
In the particular HEALPix case, the base is a rhombic dodecahedron, i.e., a polyhedron made from 12 congruent rhombic faces.\footnote{A rhombus is a quadrilateral whose four sides all have the same length.}
See \figref{pooling} for an illustration of the base rhombic dodecahedron and its subdivisions.

% HEALPix is an acronym for Hierarchical Equal Area isoLatitude Pixelization of a sphere.
The HEALPix pixelisation produces a hierarchical subdivision of a spherical surface where each pixel covers the same surface area as every other pixel.
A hierarchy is desired for the data locality in the computer memory.
Equal area is advantageous because white noise generated by the signal receiver gets integrated exactly into white noise in the pixel space.
Isolatitude is essential for the implementation of a fast spherical transform.
HEALPix is the sole pixelisation scheme which satisfies those three properties.

The lowest possible resolution is given by the base partitioning of the surface into $N_{pix} = 12$ equal-sized pixels (right-most sphere in \figref{pooling}).
As each pixel is subdivided in four, the second coarser resolution is $N_{pix} = N_{side}^2 \times 12 = 2^2 \times 12 = 48$ pixels (middle sphere in \figref{pooling}), the third is $N_{pix} = N_{side}^2 \times 12 = 4^2 \times 12 = 192$ pixels, etc., where $N_{side} = 1, 2, 4, 8, \ldots$ is the grid resolution parameter.
High-resolutions HEALPix maps easily reach millions of pixels.
% \figref{example_maps} shows two examples of HEALPix maps: the Cosmic Microwave Background \citep{planck2015overview}, galaxies found in Sloan Digital Sky Survey Data Release 14 \citep{abolfathi2017sdssDR14}, and an example simulated mass map on the footprint of Dark Energy Survey Data Release 1 \citep{des2018dr1}.

@@

% mdeff: Why not $1/d$? See Pascal's paper
% nati: The question is: is $1/d$ also good when we have 7/8 neighbors instead of 4? Practically, we can also use $1/d$, but that means changing the code and re-runing the experiment. We probably do not have to change any parameter though... @mdeff: do you think we should try this for this paper? Or we keep that for the next one.
% mdeff: I tried and the eigenvectors and eigenvalues are mostly the same. To me, there is no point in trying for trying. We should rather come up with a definition of distance that will make the eigenvectors closer to the real spherical harmonics.
% Agreed: I added a reference to Pascal paper.

@@

Finally, the graph convolution can be interpreted in the vertex domain as a scalar product with localizations $\T_i h$ of the kernel $h$. Indeed, the result of the convolution of the signal $\f$ with the kernel $h$ is
\begin{equation} \label{eqn:graph_convolution_spatial}
	(h(\L) \f)_i = \langle \T_i h(\L), \f \rangle = \langle h(\L) \b \delta^i, \f \rangle.
\end{equation}
To make the parallel with the classical 1D convolution, let $f, g\in \mathbb{Z} \rightarrow \mathbb{R}$ be two 1D discrete signals. Their convolution can be written in the same form as \eqnref{graph_convolution_spatial}:
\begin{equation*}
	(f \ast g) [i] = \sum_{j=-\infty}^\infty f[j] g[i-j] = \langle T_i g,  f \rangle,
\end{equation*}
where $T_i g[j] = g[i-j]$ is, up to a flip, a translation operator.
% nati: I think now the indexing is clear and will not confuse with the rest of the paper.
Similarly as \eqnref{graph_convolution_spatial}, the convolution of the signal $f$ by a kernel $g$ is the scalar product of $f$ with translated versions $T_i g$ of the kernel $g$.
% Signals and kernels are the same thing in classical signal processing precisely because there exists a proper definition of translation.
% For Michael: I know what you mean here, but it is slightly more complicated than that.
Additionally, it turns out that the localization operator $\T_i h$ is a generalization of the translation operator on graphs. In the particular case where the Laplacian matrix $\L$ is circulant, $\T_i h$ is a translated version of $\T_j h$ for all $i, j$ and both convolutions are equivalent. We refer the reader to \citep[Sec 2.2]{perraudin2017stationary} for a detailed discussion of the connection between translation $T_i$ and localization $\T_i$.

@@

While \eqnref{graph_convolution_fourier} is a theoretically well justified definition of the convolution of a graph signal $\f$ by a kernel $h$, it is computationally cumbersome.
As no efficient and general fast Fourier transform (FFT) exists for graphs \citep{le2018fgft}, the execution of the Fourier transform by multiplication of the signal $\f$ by the dense matrix $\U$ costs $\bO(N_{pix}^2)$ operations.
% \footnote{See \citep{le2018fgft} for a comparison between the polynomial definition of filters presented here and a class of fast graph Fourier transforms based on the approximation of the Fourier basis $\U$ by a product of sparse matrices.}
% TK: I would just cite the paper, we have not introduced the polynomials at this stage yet.
In a neural network, this operation has to be performed for each forward and backward pass.
As current training procedures require processing of many samples, it would be very slow to perform them for a quadratic computational cost in the sample size.
Moreover, the eigen-decomposition of the Laplacian $\L$ is needed to obtain the Fourier basis $\U$ in the first place.
That has a unique cost of $\bO(N_{pix}^3)$ operations.

>>>>

Convolutional Neural Networks (CNNs) are a cornerstone of the Deep Learning toolbox and have led to many breakthroughs in Artificial Intelligence.
So far, these networks have mostly been developed for regular Euclidean domains such as those supporting images, audio, or video.
Because of their success, CNN-based methods are becoming increasingly popular in Cosmology.
Cosmological data often comes as spherical maps, which make the use of traditional CNNs more complicated.
The commonly used pixelization scheme for spherical maps is the Hierarchical Equal Area isoLatitude Pixelisation (HEALPix).
We present a spherical CNN for analysis of full and partial HEALPix maps, which we call DeepSphere.
The spherical CNN is constructed by representing the sphere as a graph.
Graphs are versatile data structures that can represent pairwise relationships between objects or act as a discrete representation of a continuous manifold.
Using the graph-based representation, we define many of the standard CNN operations, such as convolution and pooling.
With filters restricted to being radial, our convolutions are equivariant to rotation on the sphere, and DeepSphere can be made invariant or equivariant to rotation.
This way, DeepSphere is a special case of a graph CNN, tailored to the HEALPix sampling of the sphere.
This approach is computationally more efficient than using spherical harmonics to perform convolutions.
We demonstrate the method on a classification problem of mass maps from two cosmological models and compare the performance of the CNN with that of two baseline classifiers, based on the power spectrum and pixel density histogram.
Our experimental results show that the performance of DeepSphere is always superior or equal to both of these baselines.
For high noise levels and for data covering only a smaller fraction of the sphere, DeepSphere achieves typically 10\% better classification accuracy than those baselines.
Finally, we show how learned filters can be visualized to introspect the neural network.
Code and examples are available at \url{https://github.com/SwissDataScienceCenter/DeepSphere}.

\end{abstract}

@@

\begin{figure*}
	\centering
	\includegraphics[width=\linewidth]{figure_architecture_v3}
	\caption{Overall network architecture, showing here three convolutional layers acting as feature extractors followed by three fully connected layers with softmax acting as the classifier.
    A convolutional layer is based on five operations: convolution, non-linearity, batch normalization, down-sampling, and pooling.
	While most operations are agnostic to the data domain, the convolution and the down-sampling have to be adapted.
	In this paper, we propose first to model the sphere with a graph and to perform the convolution on the graph.
	Graphs are versatile data structures which can model any sampling, even irregular or partial.
	Second, we propose to exploit a hierarchical pixelization of the sphere for the down-sampling operation.
	It allows the network to analyze the data at multiple scales while preserving the spatial localization of features.
	This figure shows a network that operates on the whole sphere.
	The process is the same when working with partial observations, except that the graph is only built for the region of interest.}
	\label{fig:architecture}
\end{figure*}

@@

@@

One direction to generalize CNNs to spherical data is to apply a standard 2D CNN to a grid discretisation of the sphere~\citep{boomsma2017spherical,su2017sphericalconv,coors2018spherenet}.
An alternative is to divide the sphere into small chunks and project those on flat 2D surfaces \citep{fluri2018deep,gupta2018nongaussianinformation,schmelze2017cosmologicalmodel,gillet2018deeplearning}.
While these approaches use the well-developed 2D convolution and hierarchical pooling, they are not equivariant to rotation.
Another direction is to leverage the spherical Fourier transform and to perform the convolution associated to the SO(3) rotation group in the spectral domain, thanks to the convolution theorem \citep{cohen2018sphericalcnn, esteves2017sphericalcnn}.
While the resulting convolution is equivariant to rotation, this approach is computationally expensive, even if a fast spherical Fourier transform is used.
Moreover, all those methods cannot be much accelerated when maps only span a part of the sky.

Our spherical CNN leverages convolutions on graphs and hierarchical pooling to address all three identified shortcomings: (i) rotation equivariance, (ii) computational efficiency, and (iii) partial sky observations.
The main idea is to model the discretised sphere as a graph of connected pixels: the length of the shortest path between two pixels is an approximation of the geodesic distance between them.
This approach is enabled by recent developments in graph signal processing, where fundamental tools, such as the Fourier transform, were defined for graphs \citep{shuman2013emerging}.
The graph convolution has been explored previously for omnidirectional imaging \citep{khasanova2017graphomni}, with a different down-sampling operation.
Our pooling strategy exploits a hierarchical pixelisation of the sphere to analyse the data at multiple scales.
We use the graph CNN formulation introduced in \citep{defferrard2016convolutional}, which can accommodate any sampling of the sphere.
As the Equal Area isoLatitude Pixelisation (HEALPix)~\citep{gorski2005healpix} is a popular sampling used in cosmology and astrophysics, we tailored the method to that particular sampling and call it DeepSphere.
DeepSphere is however easily used with other samplings as only two elements depend on it: (i) the choice of neighboring pixels when building the graph, and (ii) the choice of parent pixels when building the hierarchy.
The flexibility of modeling the data domain with a graph allows to easily model data that spans only a part of the sphere, or data that is not uniformly sampled.
Using a $k$-nearest neighbours graph, the convolution operation costs $\mathcal{O}(N_{pix})$ operations, where $N_{pix}$ is the number of pixels.
This is the lowest possible complexity for a convolution without approximations.
DeepSphere is readily apt to solve four tasks: (i) global classification (i.e., predict a class from a map), (ii) global regression (i.e., predict a set of parameters from a map), (iii) dense classification (i.e., predict a class for each pixel of a map), and (iv) dense regression, (i.e., predict a set of maps from a map).
Input data are spherical maps with a single value per pixel, such as the CMB temperature, or multiple values per pixel, such as surveys at multiple radio frequencies.

We give a practical demonstration of DeepSphere on cosmological model discrimination using maps of projected mass distribution on the sky \citep{chang2017curvedsky}.
This kind of maps can be created using the gravitational lensing technique \citep[see][for review]{BartelmannSchneider2001weak}.
Our maps are similar to the ones used by \citep{schmelze2017cosmologicalmodel}.
In a simplified scenario, we classify partial convergence maps into two cosmological models.
These models were designed to have very similar angular power spectrum.
We compare the performance of DeepSphere to a baseline algorithm: an SVM classifier that takes pixel histograms or power spectrum densities (PSDs) of these maps as input.
The comparison is made as a function of the additive noise level and the area of the sphere used in the analysis.
Results show that our model is always better at discriminating the maps, especially in the presence of noise.
DeepSphere is implemented with \pkg{TensorFlow} \citep{abadi2016tensorflow} and is intended to be easy to use out-of-the-box for cosmological applications.
Code and examples are available online.{\interfootnotelinepenalty=10000 \footnote{\url{https://github.com/SwissDataScienceCenter/DeepSphere/}}}

\section{Method}
\label{sec:method}

@@

Before doing any numerical analysis on the sphere, one first has to choose a tessellation, i.e., an exhaustive partition of the sphere into finite area elements, where the data under study is quantized.
The simplicity of the spherical form belies the intricacy of global analysis on the sphere.
There is no known point set that achieves the analogue of uniform sampling in Euclidean space and allows exact and invertible discrete spherical harmonic decompositions of arbitrary but band-limited functions.
Any existing proposition of practical schemes for the discrete treatment of such functions on the sphere introduces some systematic error dependent on the global properties of the point set.
While our method is applicable to any pixelisation of the sphere, two details depend on the chosen sampling: (i) the choice of neighbours in the construction of the graph, and (ii) the choice of parent nodes when coarsening the graph.
As HEALPix~\citep{gorski2005healpix} is our target application, we tailor the method to that particular sampling in the subsequent exposition.
\figref{example_maps} shows three examples of HEALPix maps: the Cosmic Microwave Background \citep{planck2015overview}, galaxies found in Sloan Digital Sky Survey Data Release 14 \citep{abolfathi2017sdssDR14}, and an example simulated mass map on the footprint of Dark Energy Survey Data Release 1 \citep{des2018dr1}.

HEALPix is a particular case of a more general class of schemes based on a hierarchical subdivision of a base polyhedron.
Another example is the geodesic grids which are based on geodesic polyhedrons, i.e., polyhedrons made of triangular faces. A counter-example is the equirectangular projection, which is not constructed from a base polyhedron, although it can be subdivided.
In the particular HEALPix case, the base is a rhombic dodecahedron, i.e., a polyhedron made from 12 congruent rhombic faces.\footnote{A rhombus is a quadrilateral whose four sides all have the same length.}
See \figref{pooling} for an illustration of the base rhombic dodecahedron and its subdivisions.

The HEALPix pixelisation produces a hierarchical subdivision of a spherical surface where each pixel covers the same surface area as every other pixel.
A hierarchy is desired for the data locality in the computer memory.
Equal area is advantageous because white noise generated by the signal receiver gets integrated exactly into white noise in the pixel space.
Isolatitude is essential for the implementation of a fast spherical transform.
HEALPix is the sole pixelisation scheme which satisfies those three properties.

The lowest possible resolution is given by the base partitioning of the surface into $N_{pix} = 12$ equal-sized pixels (right-most sphere in \figref{pooling}).
As each pixel is subdivided in four, the second coarser resolution is $N_{pix} = N_{side}^2 \times 12 = 2^2 \times 12 = 48$ pixels (middle sphere in \figref{pooling}), the third is $N_{pix} = N_{side}^2 \times 12 = 4^2 \times 12 = 192$ pixels, etc., where $N_{side} = 1, 2, 4, 8, \ldots$ is the grid resolution parameter.
High-resolutions HEALPix maps easily reach millions of pixels.

@@

@@

Finally, the graph convolution can be interpreted in the vertex domain as a scalar product with localizations $\T_i h$ of the kernel $h$. Indeed, the result of the convolution of the signal $\f$ with the kernel $h$ is
\begin{equation} \label{eqn:graph_convolution_spatial}
	(h(\L) \f)_i = \langle \T_i h(\L), \f \rangle = \langle h(\L) \b \delta^i, \f \rangle.
\end{equation}
To make the parallel with the classical 1D convolution, let $f, g\in \mathbb{Z} \rightarrow \mathbb{R}$ be two 1D discrete signals. Their convolution can be written in the same form as \eqnref{graph_convolution_spatial}:
\begin{equation*}
	(f \ast g) [i] = \sum_{j=-\infty}^\infty f[j] g[i-j] = \langle T_i g,  f \rangle,
\end{equation*}
where $T_i g[j] = g[i-j]$ is, up to a flip, a translation operator.
Similarly as \eqnref{graph_convolution_spatial}, the convolution of the signal $f$ by a kernel $g$ is the scalar product of $f$ with translated versions $T_i g$ of the kernel $g$.
Additionally, it turns out that the localization operator $\T_i h$ is a generalization of the translation operator on graphs. In the particular case where the Laplacian matrix $\L$ is circulant, $\T_i h$ is a translated version of $\T_j h$ for all $i, j$ and both convolutions are equivalent. We refer the reader to \citep[Sec 2.2]{perraudin2017stationary} for a detailed discussion of the connection between translation $T_i$ and localization $\T_i$.

@@

While \eqnref{graph_convolution_fourier} is a theoretically well justified definition of the convolution of a graph signal $\f$ by a kernel $h$, it is computationally cumbersome.
As no efficient and general fast Fourier transform (FFT) exists for graphs \citep{le2018fgft}, the execution of the Fourier transform by multiplication of the signal $\f$ by the dense matrix $\U$ costs $\bO(N_{pix}^2)$ operations.
In a neural network, this operation has to be performed for each forward and backward pass.
As current training procedures require processing of many samples, it would be very slow to perform them for a quadratic computational cost in the sample size.
Moreover, the eigen-decomposition of the Laplacian $\L$ is needed to obtain the Fourier basis $\U$ in the first place.
That has a unique cost of $\bO(N_{pix}^3)$ operations.

Fortunately, both of these computational issues are overcome by defining the convolution kernel $h$ as a polynomial $h_\theta(\lambda) = \sum_{k=0}^K \theta_k \lambda^k$ of degree $K$ parametrised by $K+1$ coefficients $\theta$.
The filtering operation \eqnref{graph_convolution_fourier} becomes
\begin{equation} \label{eqn:graph_convolution_monomial}
	h_\theta(\L) \f =  \U \left(\sum_{k=0}^K \theta_k \bLambda^k \right) \U\trans \f = \sum_{k=0}^K \theta_k \L^k \f,
\end{equation}
where $\L^k$ captures $k$-neighborhoods.\footnote{The 1-neighborhood of a node is the set of nodes that are directly connected to it. The $k$-neighborhood of a node is the set of nodes that are connected to it through paths of length $k$.}
The entry $(\L^k)_{ij}$ is the sum of all weighted paths of length $k$ between nodes $v_i$ and $v_j$, where the weight of a path is the multiplication of all the edge weights on the path.
Hence, it is non-zero if and only if nodes $v_i$ and $v_j$ are connected by at least one path of length $k$.
Filtering with a polynomial convolution kernel can thus be interpreted in the pixel (vertex) domain as a weighted linear combination of neighboring pixel values.
In the classical setting, convolutions are also weighted local sums.
The weights are however given by the filter coefficients only, and there is one coefficient per pixel in the patch.
In the graph setting, the weights are determined by the filter coefficients $\theta$ and the Laplacian $\L$, and there is one coefficient per neighborhood, not node.
That defines radial filters, values of which depend only on the distance to the center, and not on the direction.
It is required as there is no way to order and match nodes from the same neighborhood in a general graph.
The number of nodes in a neighborhood most often even vary from node to node.
While it may seem odd to restrict filters to be 1D while the sphere is 2D, radial filters are direction-less and result in rotation equivariant convolutions.
We note that restricting the graph convolutional kernel to a polynomial is similar to restricting the classical euclidean convolution to a fixed-size patch.
Similarly, each line of the matrix $\sum_{k=0}^K \theta_k \L^k$ defines an irregular patch of radius $K$.
Hence, filters designed as polynomials of the Laplacian have local support in the vertex domain.
